Impacts of storm and severe weather events in the United States
================================================================================ 
## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project uses the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to study the impacts of major storms and weather events in the United States in population health and economy. This project can also be used to find the weather events that cause the highest number of casualties. It can also help in assessment of the property and crop damage during the weather events. 

## Data Processing
The data was retrieved from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. 
```{r download, eval = FALSE}
setwd("./StormData")
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists()){
  download.file(url,"StormData.bz2")
  unzip("StormData.bz2")
}
```

The Storm Data file is in a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

- National Weather Service Storm Data [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
- National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Exploring the data
```{r}
stormData<- read.csv("StormData.bz2")
head(stormData)
```
The dimension of the data is given by:
  ```{r}
dim(stormData)
```
